Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-04-04 manuscript.tex
continue M&M.

Methods details below. All human iPSC and ESC experiments were approved by the Ethical Committee of the KU Leuven (%TODO Number%)  

@@

\subsubsection*{Chicken embryos}
Fertilized chicken eggs, obtained from ... %TODO
were incubated at \SI{37}{\degreeCelsius} at \SI{70}{\percent} humidity for 30 hours to reach Humburger Hamilton (HH) stage 4. Afterwards, embryos were removed from the egg and kept in culture for an additional 24 hours.

@@

\paragraph*{10x multiome library preparation}
%TODO Valerie

\paragraph*{10x multiome library sequencing}
%TODO Valerie

@@

\paragraph*{10x multiome library preparation}
%TODO Camiel

\paragraph*{10x multiome library sequencing}
%TODO Camiel

@@

\paragraph*{Samples}
Eight individual single-cell multiome libraries were generated:
\begin{itemize}
  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{1.5}{\percent} PEG hydrogel matrix.
  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix.
  \item Two libraries containing a mix of organoids at 1, 2, 3, 6 and 7 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix.
  \item A single library containing a mix of organoids at 6 and 7 days of differentiation cultured in \SI{2}{\percent}
    PEG hydrogel matrix.
  \item A single library containing a mix of organoids at 4, 5, 8 and 9 days of differentiation cultured in
    \SI{2}{\percent} PEG hydrogel matrix.
  \item A single library containing a mix of organoids at 8 and 9 days of differentiation cultured in \SI{2}{\percent}
    PEG hydrogel matrix.
  \item A single library containing a mix of organoids at 10 and 11 days of differentiation cultured in \SI{2}{\percent}
    PEG hydrogel matrix.
\end{itemize}

@@

\paragraph*{Cell line demultiplexing}
Variants were called from in house Omni-ATAC-seq\cite{corcesImprovedATACseqProtocol2017} data that was performed on each
cell line individually (unpublished data). 
First, bam files per cell line were merged using samtools (v1.11; \href{https://rrid.site/data/record/nlx_144509-1/SCR_002105/resolver}{RRID:SCR\_002105}) merge, sorted using samtools sort and indexed using samtools index. 
Next, for each cell line merged bam file common variants were called using the bcftools (v1.11; \href{https://rrid.site/data/record/nlx_144509-1/SCR_005227/resolver}{RRID:SCR\_005227}) mpileup command using common short genetic variants from 1000 Genomes project as targets parameter (-T) and 8,000 as max depth (-D) parameter followed by the call command from bcftools.

next ... demuxlet etc

@@

\paragraph*{single-cell ATAC quality control}

>>>>

Methods details below. All human iPSC and ESC experiments were approved by the Ethical Committee of the KU Leuven  Number
%TODO: add ECD

@@

\subsubsection*{Chicken embryos}
Fertilized chicken eggs, obtained from ... 
%TODO: add vendor
were incubated at \SI{37}{\degreeCelsius} at \SI{70}{\percent} humidity for 30 hours to reach Humburger Hamilton (HH) stage 4. Afterwards, embryos were removed from the egg and kept in culture for an additional 24 hours.

@@

\paragraph*{10x multiome library preparation}
%TODO: Valerie

\paragraph*{10x multiome library sequencing}
%TODO: Valerie

@@

\paragraph*{10x multiome library preparation}
%TODO :Camiel

\paragraph*{10x multiome library sequencing}
%TODO: Camiel

@@

\paragraph*{Samples}
Eight individual single-cell multiome libraries were generated:
\begin{itemize}
  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{1.5}{\percent} PEG hydrogel matrix (IPS\_\_081787\_\_575955\_\_10x\_multiome\_neural\_tube\_organoids\_1\_5\_PEG).
  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix (IPS\_\_c13535\_\_a58829\_\_10x\_multiome\_neural\_tube\_organoids\_2\_PEG).
  \item Two libraries containing a mix of organoids at 1, 2, 3, 6 and 7 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix (6-7\_1-2-3\_1 and 6-7\_1-2-3\_2).
  \item A single library containing a mix of organoids at 6 and 7 days of differentiation cultured in \SI{2}{\percent}
    PEG hydrogel matrix (6-7).
  \item A single library containing a mix of organoids at 4, 5, 8 and 9 days of differentiation cultured in
    \SI{2}{\percent} PEG hydrogel matrix (8-9\_4-5).
  \item A single library containing a mix of organoids at 8 and 9 days of differentiation cultured in \SI{2}{\percent}
    PEG hydrogel matrix (8-9).
  \item A single library containing a mix of organoids at 10 and 11 days of differentiation cultured in \SI{2}{\percent}
    PEG hydrogel matrix (10-11).
\end{itemize}

@@

\paragraph*{Cell line demultiplexing}
Variants were called from in house Omni-ATAC-seq\cite{corcesImprovedATACseqProtocol2017} data that was performed on each
cell line individually (unpublished data). 
First, bam files per cell line were merged using samtools\cite{liSequenceAlignmentMap2009} (v1.11; \href{https://rrid.site/data/record/nlx_144509-1/SCR_002105/resolver}{RRID:SCR\_002105}) merge, sorted using samtools sort and indexed using samtools index. 
Next, for each cell line merged bam file common variants were called using the bcftools\cite{liStatisticalFrameworkSNP2011} (v1.11; \href{https://rrid.site/data/record/nlx_144509-1/SCR_005227/resolver}{RRID:SCR\_005227}) mpileup command using common short genetic variants from 1000 Genomes project as targets parameter (-T) and 8,000 as max depth (-D) parameter followed by the call command from bcftools. 
Next, single-cell multiome samples were demultixpled per cell line using popscle (\href{https://rrid.site/data/record/nlx_144509-1/SCR_026707/resolver}{RRID:SCR\_026707}) demuxlet\cite{kangMultiplexedDropletSinglecell2018} after first sorting the variant VCF file as the single-cell multiome sample bam file, using a custom script sort\_VCF\_same\_as\_bam, filtering the sample bam file to only retain reads that overlap with SNPs in the VCF file and have a valid barcode passing the Cell Ranger ARC quality control, using a custom script filter\_bam\_file\_for\_popscle\_dsc\_pileup, (custom scripts can be found in the popscle\_helper\_tools GitHub repository: \href{https://github.com/aertslab/popscle_helper_tools}{https://github.com/aertslab/popscle\_helper\_tools}) and creating a pileup of reads for each SNP in the VCF file and each cell barcode using the single-cell multiome sample bam file and finally popscle demuxlet was run on the sample pileup, variant VCF file using GT as field (--field) parameter.

\paragraph*{single-cell RNA quality control}
Cells with less than 3 genes expressed and genes expressed in less than 200 cells were removed using the Scanpy\cite{wolfSCANPYLargescaleSinglecell2018} (v1.9.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) functions filter\_cells and filter\_genes.
Next, quality control metrics were calculated using the Scanpy function calculate\_qc\_metrics using qc\_vars=["mt"],
percent\_top = None and log1p = False as parameters.
Cells with less than \SI{25}{\percent} mitochondrial counts, more than a total of 2,000 counts and less than a total of
9,000 genes with counts were kept resulting in a total of 27,552 cells passing quality control.

\paragraph*{single-cell RNA batch effect correction}
Batch effect correction was performed using harmonypy\cite{korsunskyFastSensitiveAccurate2019} (v0.0.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_022798/resolver}{RRID:SCR\_022798}) using as batch key the combination of experimental batch and cell line on 50 PCA coordinates calculated using the Scanpy \cite{wolfSCANPYLargescaleSinglecell2018} (v1.9.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) function pca.

\paragraph*{single-cell RNA cell type annotation}
Cells were first annotated as Facial mesenchyme, Neural crest, Neuron, Peripheral neuron, Pluripotent stem cell or
Neuronal progenitor by 
first normalizing, log transforming and scaling the gene expression matrix using the Scanpy\cite{wolfSCANPYLargescaleSinglecell2018} (v1.9.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) functions normalize\_per\_cell, log1p and scale
and clustering the cells using the Leiden algorithm\cite{traagLouvainLeidenGuaranteeing2019} at a resolution of 5.1 (making use of the leiden function from Scanpy), 
calculating the median expression per cluster, 
binarizing this matrix using a gene expression threshold of 0, 
and calculating the jaccard distance using Scipy\cite{virtanenSciPy10Fundamental2020} (v1.9.3; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}) between the binarized median expression matrix and a knowledge matrix where containing marker genes per cell type and where the value is 1 for markers of that cell type and 0 otherwise constructed from the marker gene table (\textbf{Supplemental table S1}). 
For each cluster, the annotation with minimal distance was selected breaking ties by selecting a random annotation with equal distance.
Neuronal progenitors were further annotated as Neuronal progenitor, Neuronal progenitor early neuron, KRT19+ neuronal
progenitors, FOXG1+ neuronal progenitors, NFE2L3+ neuronal progenitors and KRT19+/NFE2L3+ neuronal progenitors using the
same procedure at a Leiden resolution of 1.9 after subsetting for neuronal progenitors and re-normalizing, scaling and
integrating (see batch effect correction) the data.
non FOXG1+, NFE2L3+ or KRT19+ neuronal progenitors were further annotated as FP, p3, p0-2, dp1-6 or RP by subsetting non FOXG1+, NFE2L3+ or KRT19+ neuronal progenitors, re-normalizing, scaling and integrating (see batch effect correction) the data. Next, a force-directed graph embedding\cite{islamCharacterizationSinglecellTranscriptional2011, jacomyForceAtlas2ContinuousGraph2014} was calculated using the Scanpy function draw\_graph followed by diffusion pseudotime inference\cite{haghverdiDiffusionPseudotimeRobustly2016} using the Scanpy function dpt and binned in 15 and 50 bins using the Pandas\cite{thepandasdevelopmentPandasdevPandasPandas2024} (v1.5.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018214/resolver}{RRID:SCR\_018214}) cut function. 
Using the same procedure as described above cell type annotations were determined for binned pseudotime in 50 bins and
for each bin of the binned pseudotime in 15 bins the most common annotation was selected.
Neural crest cells were further annotated as pre-migratory neural crest, bonafide neural crest, EMT neural crest,
mesenchyme and peripheral neuron manually based on the expression of marker genes.
Neuron cells and Neuronal progenitor early neuron cells were further annotated as ASCL1+ Neurons, SST+ Neurons, SOX2+ Neurons,
ASCL1+ Neurons, UNCX+ Neurons, GAD1+ Neurons and GATA2+ Neurons based on the expression of marker genes.

\paragraph*{single-cell ATAC consensus peaks}
A set of consensus peaks was called by first generating pseudobulk ATAC-seq profiles for each of the highest resolution
cell types (see single-cell RNA cell type annotation) using the export\_pseudobulk function, with default parameters, from pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023}
(v1.0.3.dev20+g8955c76, \href{}{RRID:SCR\_026618}) followed by peak calling using MACS\cite{zhangModelbasedAnalysisChIPSeq2008} (v2.2.9.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_013291/resolver}{RRID:SCR\_013291} ) using the pycisTopic function peak\_calling, using default parameters, and consensus peak generation using the pycisTopic function get\_consensus\_peaks, using a peak width of 500 bp and default parameters. This resulted in 488,211 consensus peaks.

\paragraph*{single-cell ATAC quality control}
Sample and barcode level quality control data was calculated using the compute\_qc\_stats function from
pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, \href{}{RRID:SCR\_026618})
cells were filtered based on the number of unique fragments per cell, Fraction of Reads in Peaks (FRIP) and
transcription start site (TSS) enrichment. Following thresholds were used for each sample:
\begin{itemize}
  \item sample IPS\_\_081787\_\_575955\_\_10x\_multiome\_neural\_tube\_organoids\_1\_5\_PEG
    \begin{itemize}
      \item log unique number of fragments: \textgreater 3.5
      \item FRIP: \textgreater 0.5
      \item TSS enrichment: \textgreater 5
    \end{itemize}
  \item sample IPS\_\_c13535\_\_a58829\_\_10x\_multiome\_neural\_tube\_organoids\_2\_PEG
  \begin{itemize}
      \item log unique number of fragments: \textgreater 3.5
      \item FRIP: \textgreater 0.5
      \item TSS enrichment: \textgreater 5
    \end{itemize}
  \item sample 10-11
  \begin{itemize}
      \item log unique number of fragments: \textgreater 3.25
      \item FRIP: \textgreater 0.45
      \item TSS enrichment: \textgreater 6
    \end{itemize}
  \item sample 6-7
  \begin{itemize}
      \item log unique number of fragments: \textgreater 3.5
      \item FRIP: \textgreater 0.45
      \item TSS enrichment: \textgreater 6
    \end{itemize}
  \item sample 6-7\_1-2-3\_1
  \begin{itemize}
      \item log unique number of fragments: \textgreater 3.75
      \item FRIP: \textgreater 0.5
      \item TSS enrichment: \textgreater 7
    \end{itemize}
  \item sample 6-7\_1-2-3\_2
  \begin{itemize}
      \item log unique number of fragments: \textgreater 3.75
      \item FRIP: \textgreater 0.5
      \item TSS enrichment: \textgreater 6
    \end{itemize}
  \item sample 8-9
  \begin{itemize}
      \item log unique number of fragments: \textgreater 3.25
      \item FRIP: \textgreater 0.4
      \item TSS enrichment: \textgreater 7
    \end{itemize}
  \item sample 8-9\_4-5
  \begin{itemize}
      \item log unique number of fragments: \textgreater 3.75
      \item FRIP: \textgreater 0.5
      \item TSS enrichment: \textgreater 6
    \end{itemize}
\end{itemize}

Cells that were predicted as doublets by demuxlet (see: Cell line demultiplexing) were also filtered out resulting in a
total of 28,705 cells passing quality control.

\paragraph*{single-cell ATAC topic modeling}
Four topic models were generated: one on all cells, one on the subset of cells annotated as neuronal progenitors, one on
the subset of cells annotated as neuron (Neuron cells and Neuronal progenitor early neuron cells) and one on the subset of cells annotated as neural crest.
Topic modeling was performed using Mallet\cite{mccallumMALLETMachineLearning} (v202108; RRID:SCR\_026708) using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID:SCR\_026618}) function run\_cgs\_models\_mallet using 500 iterations, alpha equal to 50, eta equal to 0.1, alpha\_by\_topic equal to True and eta\_by\_topic equal to False and using random seed 555. 
Topic modeling was performed separately for all cells and for each subset of cells.
Model evaluation was performed using the pycisTopic function evaluate\_model that calculates model coherence\cite{mimno-etal-2011-optimizing}, log-likelihood\cite{griffithsFindingScientificTopics2004}, model density\cite{arunFindingNaturalNumber2010} and a divergence-based metric\cite{caoDensitybasedMethodAdaptive2009}.
A model of 50 topics was selected for the model containing all cells,
a model of 30 topics was selected for the model containing neuronal progenitors,
a model of 25 topics was selected for the model containing neurons and
a model of 20 topics for the model containing neural crest cells.
The region topic probability matrix was thresholded using the binarize\_topics function from pycisTopic using the otsu
method\cite{otsuThresholdSelectionMethod1979} and the cell topic probability matris using the same function using the li
method\cite{liMinimumCrossEntropy1993}.

\paragraph*{Training $DeepNeuralTube_O$}
A deepTopic model\cite{minnoyeCrossspeciesAnalysisEnhancer2020}, now implemented in
CREsted\cite{kempynckCREstedModelingGenomic2025} (\href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}), containing 5 convolutional and max pooling layers: dimensions (500,
1,024), (125, 512), (32, 512), (8, 512) and (2, 512) a fully connected dense layer of dimension 1,024 and a dense output
layer of dimension 75 was trained to classify consensus peaks in topics based on their genomic sequence. GELU was used as activaiton function, 17 as
kernel size, 4 as max pooling size, a dropout of 0.15 was used for the convolutional layers, a dropout of 0.5 was used
for the dense layers, the convolutional layers were l2 normalized using an epsilon value of 1e-6 and the dense layers were
l2 normalized using an epsilon value of 1e-3. The Adam optimizer was used with a learning rate of 0.001 that was dynamically scaled and binary cross entropy as a loss function and a batch size of 1,024.
Model was built and trained using Tensorflow (v2.41; \href{https://rrid.site/data/record/nlx_144509-1/SCR_016345/resolver}{RRID:SCR\_016345}). 
As training data all binarized topics (see single-cell ATAC topic modeling) from the subset of neuronal progenitors,
neurons and neural crest together with 10 topics form the topic model containing all cells representing the pluripotent
stem cell state were used for a total of 75 classes/topics. 
Data was augmented by training on both forward and reverse complement sequence and \SI{10}{\percent} of the data was
used for validation and \SI{10}{\percent} for testing for a total of 5,192,175 training samples, 650,280 test samples
and 647,805 validation samples. 
Based on the validation area under the precision recall curve and validation area under the receiver operator curve the
model after 23 epochs was selected.
The classes of the model correspond as follows to the different data subsets: 
\begin{itemize}
  \item $0 \leq class_{index} < 25 $: neuron
  \item $25 \leq class_{index} < 55$: Neuronal progenitor
  \item $55 \leq class_{index} < 65$: Neural crest
  \item $65 \leq class_{index} < 75$: pluripotent stem cell
\end{itemize}

\subsubsection*{Human embryo data analysis}

\paragraph*{RNA count matrices and ATAC fragment files}

% TODO: Camiel

@@

% TODO: Camiel

\paragraph*{single-cell RNA cell type annotation}

%TODO: Camiel

\paragraph*{single-cell ATAC consensus peaks}

%TODO: Camiel

\paragraph*{single-cell ATAC quality control}
Sample and barcode level quality control data was calculated using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v2.0a, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID:SCR\_026618}) command pycistopic qc, using a TSS annotation bed file that was generated using the pycisTopic command pycistopic tss get\_tss using hsapiens\_gene\_ensembl as biomart name. High quality cells were selected based on the number of unique fragments per cell and TSS enrichment. The minimal number of unique fragments per cell was automatically determined using otsu thresholding \cite{otsuThresholdSelectionMethod1979} and a TSS enrichment threshold of 7 was used. This resulted in 38,998 high quality cells passing quality control.

\paragraph*{single-cell ATAC topic modeling}
Three topic models were generated: one on the subset of cells annotated as neuronal progenitors, on on the subset of
cells annotated as neuron and one on the subset of cells annotated as neural crest. First, a fragment count matrix was
generated for each sample using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v2.0a, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID:SCR\_026618}) function create\_fragments\_matrix\_from\_fragments, this matrix was binarized (\textgreater 0), merged in one count matrix using the Scipy\cite{virtanenSciPy10Fundamental2020} (v1.15.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}) function sparse.hstack and subsequently split in a matrix for each of neuronal progenitors, neurons and neural crest cells and saved as matrix market format. A mallet corpus file was generated for each binary matrix using the pycisTopic command pycistopic topic\_modeling mallet create\_corpus and finally topic modeling was run using Mallet\cite{mccallumMALLETMachineLearning} (v202108; RRID:SCR\_026708) using the pycisTopic command: pycistopic topic\_modeling mallet run using 300 iterations and default parameters otherwise. Model evaluation was performed using the pycisTopic command pycistopic topic\_modeling mallet stats that calculates model coherence\cite{mimno-etal-2011-optimizing}, log-likelihood\cite{griffithsFindingScientificTopics2004}, model density\cite{arunFindingNaturalNumber2010} and a divergence-based metric\cite{caoDensitybasedMethodAdaptive2009}.
A model of 60 topics was selected for the model containing neuronal progenitors,
a model of 30 topics was selected for the model containing neurons and
a model of 30 topics for the model containing neural crest cells.
The region topic probability matrix was thresholded using the binarize\_topics function from pycisTopic using the otsu
method\cite{otsuThresholdSelectionMethod1979} and the cell topic probability matris using the same function using the li
method\cite{liMinimumCrossEntropy1993}.

\paragraph*{Training $DeepNeuralTube_E$}
A deepTopic model\cite{minnoyeCrossspeciesAnalysisEnhancer2020}, now implemented in
CREsted\cite{kempynckCREstedModelingGenomic2025} (\href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}), containing 5 convolutional and max pooling layers: dimensions (500,
1,024), (125, 512), (32, 512), (8, 512) and (2, 512) a fully connected dense layer of dimension 1,024 and a dense output
layer of dimension 75 was trained to classify consensus peaks in topics based on their genomic sequence. GELU was used as activaiton function, 17 as
kernel size, 4 as max pooling size, a dropout of 0.15 was used for the convolutional layers, a dropout of 0.5 was used
for the dense layers, the convolutional layers were l2 normalized using an epsilon value of 1e-6 and the dense layers were
l2 normalized using an epsilon value of 1e-3. The Adam optimizer was used with a learning rate of 0.001 that was dynamically scaled and binary cross entropy as a loss function and a batch size of 1,024.
Model was built and trained using Tensorflow (v2.41; \href{https://rrid.site/data/record/nlx_144509-1/SCR_016345/resolver}{RRID:SCR\_016345}). 
As training data all binarized topics (see single-cell ATAC topic modeling) from the subset of neuronal progenitors,
neurons and neural crest together were used for a total of 120 classes/topics. 
Data was augmented by training on both forward and reverse complement sequence and \SI{10}{\percent} of the data was
used for validation and \SI{10}{\percent} for testing for a total of 11,935,075 training samples, 1,494,835 test samples
and 1,497,350 validation samples. 
Based on the validation area under the precision recall curve and validation area under the receiver operator curve the
model after 36 epochs was selected.
The classes of the model correspond as follows to the different data subsets: 
\begin{itemize}
  \item $0 \leq class_{index} < 30 $: neuron
  \item $30 \leq class_{index} < 90$: Neuronal progenitor
  \item $90 \leq class_{index} < 120$: Neural crest
\end{itemize}

\subsubsection*{Cell cycle annotation}
Marker genes of the S-phase and G2 and M phase of the cell cycle\cite{tiroshDissectingMulticellularEcosystem2016} were
used in the Scanpy\cite{wolfSCANPYLargescaleSinglecell2018} (v1.11.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) function score\_genes\_cell\_cycle.

\subsubsection*{Motif enrichment analysis}

A cistarget motif ranking database was created by first getting the genomic sequence of each organoid consensus peak with 1,000 bp background padding using the create\_fasta\_with\_padded\_bg\_from\_bed script from the create\_cisTarget\_databases repository (\href{https://github.com/aertslab/create_cisTarget_databases}{https://github.com/aertslab/create\_cisTarget\_databases}) and scoring each DNA sequence using Cluster-Buster\cite{frithClusterBusterFindingDense2003}(\href{https://rrid.site/data/record/nlx_144509-1/SCR_024810/resolver?}{RRID:SCR\_024810}) using the create\_cisTarget\_databases command create\_cistarget\_motif\_databases followed by the generation of the ranking databases using the command convert\_motifs\_or\_tracks\_vs\_regions\_or\_genes\_scores\_to\_rankings\_cistarget\_dbs. Next, the SCENIC+\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0a2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026702/resolver?}{RRID:SCR\_026702}) command scenicplus grn\_inference motif\_enrichment\_cistarget command was used to perform cisTarget\cite{potierUsingCisTargetXPredict2012} motif enrichment analysis for each organoid and embryo topic.

\subsubsection*{Annotating embryo cells based on organoid topics}
A matrix multiplication between the region-topic probabilities of organoid topics and the binarized embryo fragment
count matrix was performed to obtain an imputed embryo cell topics probability matrix (where the topics are organoid
topics). This matrix was binarized by taking the top 1,000 cells per topic.

\subsubsection*{\textit{de novo} motif discovery}
For all organoid and embryo topics using respectively $DeepNeuralTube_O$ and $DeepNeuralTube_E$ attribution scores
were calculated using the CREsted\cite{kempynckCREstedModelingGenomic2025} (v1.3.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}) function integrated\_grad, on 1,000 regions with the highest prediction score for that topic according to the model. Next, patterns/motifs were discovered \textit{de novo} using a custom implementation of tfmodisco-lite (v2.0.0; RRID:SCR\_026709). First seqlets were called using the tfmodisco-lite function extract\_seqlets using the attribution scores as input and using parameters 
sliding\_window\_size = 15, 
flank = 5, 
suppress = 12, 
target\_fdr = 0.2, 
min\_passing\_windows\_frac = 0.03, 
max\_passing\_windows\_frac = 0.3 and 
weak\_threshold\_for\_counting\_sig = 0.8. 
Next, the attribution score of each seqlet was extracted based on the seqlet coordinates and negative and positive seqlets were defined based on the sum of the attribution score without the flank filtering out seqlets for which the abolute value of the sum of the attribution score was lower than or equal to X. 
Finally, separately for positive and negative seqlets 1,000 seqlets were sampled and used in the tfmodisco-lite function seqlet\_to\_patterns using parameters 
min\_overlap\_while\_sliding = 0.7,
nearest\_neighbors\_to\_compute = 500,
affmat\_correlation\_threshold = 0.15,
tsne\_perplexity = 10,
n\_leiden\_iterations = -1,
n\_leiden\_runs = 50,
frac\_support\_to\_trim\_to = 0.2,
min\_num\_to\_trim\_to = 30,
trim\_to\_window\_size = 20,
initial\_flank\_to\_add = 5,
prob\_and\_pertrack\_sim\_merge\_thresholds = [(0.8,0.8), (0.5, 0.85), (0.2, 0.9)],
prob\_and\_pertrack\_sim\_dealbreaker\_thresholds = [(0.4, 0.75), (0.2,0.8), (0.1, 0.85), (0.0,0.9)],
subcluster\_perplexity = 50,
merging\_max\_seqlets\_subsample = 300,
final\_min\_cluster\_size = 20,
min\_ic\_in\_window = 0.6,
min\_ic\_windowsize = 6 and 
ppm\_pseudocount= 0.001.
Next, similar patterns were clustered together by first calculating a pairwise pattern-to-pattern similarity metric
using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}). The Pearson correlation of one minus the E-value (product of the TomTom p-value and the number of patterns), calculated using the corrcoef function from NumPy\cite{harrisArrayProgrammingNumPy2020} (v2.0.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008633/resolver}{RRID:SCR\_008633}), was used for hierarchical clustering using the hierarchy.linkage (using average as method) and fcluster (using maxclust as criterion and a manual defined number of clusters) functions from Scipy\cite{virtanenSciPy10Fundamental2020} (v1.15.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}).

\subsubsection*{Validation of DeepNeuralTube organoid based on Vista enhancers}
Data was downloaded from the VISTA Enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025} (\href{https://gitlab.com/egsb-mfgl/vista-data/-/raw/main/experiments.tsv.gz}{https://gitlab.com/egsb-mfgl/vista-data/-/raw/main/experiments.tsv.gz}) and the DNA sequence of each element was obtain either from the human genome (hg38) or the mouse genome (mm10) dependent on the origin of each sequence. In case a sequence was shorter than 500 bp (the input size of DeepNeuralTube) the sequence was extended to 500 bp based on the genomic coordinates. In case the sequence was larger than 500 bp a sliding window, with step size of 1, was produced over the sequence. A prediction score for each sequence and using both $DeepNeuralTube_O$ and $DeepNeuralTube_E$ was calculated and per experiment and output class of the model the maximum prediction score was retained. 
Ground truth neural tube enhancers were defined as enhancers that were positive in the tissue "nt", in at least 3 embryos and at
least \SI{50}{\percent} of tested embryos. 
Ground truth facial mesenchyme enhancers were defined as enhancers that were positive in the tissue "fm", in at least 3 embryos and at
least \SI{50}{\percent} of tested embryos. 
For both neural tube enhancers and facial mesenchyme enhancers precision-recall curves and receiver-operator curves were
calculated using respectively the function metrics.precision\_recall\_curve and metrics.roc\_curve from Scikit-learn
(v1.5.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_002577/resolver}{RRID:SCR\_002577}) using the maximum
prediction score across classes related to neural tube respectively facial mesenchyme in both organoid and embryo. 
We consider classes related to neural tube in the organoid as classes of the following topics: 
neural crest Topic 5,
neural crest Topic 7,
neuron Topic 1,
neuron Topic 2,
neuron Topic 3,
neuron Topic 4,
neuron Topic 6,
neuron Topic 10,
neuron Topic 11,
neuron Topic 12,
neuron Topic 13,
neuron Topic 15,
neuron Topic 16,
neuron Topic 18,
neuron Topic 19,
neuron Topic 20,
neuron Topic 21,
neuron Topic 23,
neuron Topic 24,
neuron Topic 25,
progenitor Topic 1,
progenitor Topic 3,
progenitor Topic 8,
progenitor Topic 9,
progenitor Topic 11,
progenitor Topic 13,
progenitor Topic 14,
progenitor Topic 16,
progenitor Topic 19,
progenitor Topic 21,
progenitor Topic 23,
progenitor Topic 24,
progenitor Topic 25,
progenitor Topic 29 and
progenitor Topic 30
We consider classes related to neural tube in the embryo as classes of the following topics: 
│   │   "neuron_Topic_1",  #
     7   │   │   "neuron_Topic_3",  #
     6   │   │   "neuron_Topic_5",  #
     5   │   │   "neuron_Topic_6",  #
     4   │   │   "neuron_Topic_7",
     3   │   │   "neuron_Topic_8",  #
     2   │   │   "neuron_Topic_9",  #
     1   │   │   "neuron_Topic_10",
  1016   │   │   "neuron_Topic_11",  #
     1   │   │   "neuron_Topic_12",
     2   │   │   "neuron_Topic_13",  #
     3   │   │   "neuron_Topic_14",
     4   │   │   "neuron_Topic_15",  #
     5   │   │   "neuron_Topic_17",  #
     6   │   │   "neuron_Topic_18",  #
     7   │   │   "neuron_Topic_19",  #
     8   │   │   "neuron_Topic_22",  #
     9   │   │   "neuron_Topic_24",  #
    10   │   │   "neuron_Topic_26",  #
    11   │   │   "neuron_Topic_27",  #
    12   │   │   "neuron_Topic_29",
    13   │   │   "neuron_Topic_30",
    14   │   │   "progenitor_Topic_1",  #
    15   │   │   "progenitor_Topic_3",
    16   │   │   "progenitor_Topic_4",  #
    17   │   │   "progenitor_Topic_8",  #
    18   │   │   "progenitor_Topic_10",
    19   │   │   "progenitor_Topic_11",
    20   │   │   "progenitor_Topic_16",
    21   │   │   "progenitor_Topic_17",
    22   │   │   "progenitor_Topic_21",  #
    23   │   │   "progenitor_Topic_22",  #
    24   │   │   "progenitor_Topic_28",  #
    25   │   │   "progenitor_Topic_29",  #
    26   │   │   "progenitor_Topic_31",  #
    27   │   │   "progenitor_Topic_32",  #
    28   │   │   "progenitor_Topic_36",
    29   │   │   "progenitor_Topic_40",  #
    30   │   │   "progenitor_Topic_41",  #
    31   │   │   "progenitor_Topic_44",  #
    32   │   │   "progenitor_Topic_45",
    33   │   │   "progenitor_Topic_49",  #
    34   │   │   "progenitor_Topic_51",
    35   │   │   "progenitor_Topic_57",  #
    36   │   │   "progenitor_Topic_58",  #
    37   │   │   "progenitor_Topic_59",
    38   │   │   "neural_crest_Topic_12",  #
    39   │   │   "neural_crest_Topic_13",  #
    40   │   │   "neural_crest_Topic_15",
    41   
We consider classes related to facial mesenchyme in the organod as classes of the following topics: 
neural crest Topic 3
We consider classes related to facial mesenchyme in the organod as classes of the following topics: 
neural crest Topic 1, 
neural crest Topic 19, 
neural crest Topic 21 and 
neural crest Topic 29
